---
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: false
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Finding overlaps between GenomicRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}  
---

# Finding overlaps between *GRanges* objects

Interval overlapping is the process of comparing the ranges in two
objects to determine if and when they overlap. As such, it is perhaps
the most common operation performed on *GRanges* objects. 
To this end, the `r BiocStyle::Biocpkg("GenomicRanges")`
package provides a family of interval overlap functions. The most general
of these functions is `findOverlaps()`, which takes a query and a
subject as inputs and returns a *Hits* object containing
the index pairings for the overlapping elements.

Let us assume that we have three random data.frame objects, each with
annoyingly differing ways of naming the columns defining the ranges,
```{r reps}
set.seed(66+105+111+99+49+56)

pos <- sample(1:200, size = 30L)
size <- 10L
end <- size + pos - 1L
chrom <- sample(paste0("chr", 1:3), size = 30L, replace = TRUE)
query_df <- data.frame(chrom = chrom, 
                       start = pos,
                       end = end)
query_dfs <- split(query_df, 1:3)
q1 <- rename(query_dfs[[1L]], start = "pos")
q2 <- rename(query_dfs[[2L]], chrom = "ch", start = "st")
q3 <- rename(query_dfs[[3L]], end = "last")
```
The `makeGRangesFromDataFrame()` function can guess some of these, but
not all of them, so we help it out,
```{r makeGRangesFromDataFrame}
q1 <- makeGRangesFromDataFrame(q1, start.field = "pos")
q2 <- makeGRangesFromDataFrame(q2, seqnames.field = "ch",
                                 start.field = "st")
q3 <- makeGRangesFromDataFrame(q3, end.field = "last")
query <- mstack(q1, q2, q3, .index.var="replicate")
sort(query, by = ~ start)
```
Above, we use the convenient `mstack()` function, which stacks its
arguments, populating the `.index.var=` column with the origin of each
range (using the argument names or positions).

Perhaps the simplest overlap-based operation is `subsetByOverlaps()`,
which extracts the elements in the query (the first argument) that
overlap at least one element in the subject (the second).

```{r subsetByOverlaps}
subject <- gr
subsetByOverlaps(query, subject, ignore.strand=TRUE)
```
In every call to an overlap operation, it is necessary to specify
`ignore.strand=TRUE`, except in rare cases when we do not want ranges
on opposite strands to be considered overlapping.

To generally compute on the overlaps, we call `findOverlaps()` to
return a `Hits` object, which is essentially a bipartite graph
matching query ranges to overlapping subject ranges.

```{r findOverlaps}
hits <- findOverlaps(query, subject, ignore.strand=TRUE)
```

We typically use the hits to perform one of two operations: join and
aggregate. For example, we could inner join the scores from the
subject using the query and subject indexes,
```{r innerJoin}
joined <- query[queryHits(hits)]
joined$score <- subject$score[subjectHits(hits)]
```
The above carries over a single metadata column from the
subject. Similar code would carry over other columns and even the
ranges themselves. 

Sometimes, we want to merge the matched query and subject ranges,
typically by finding their intersection,
```{r overlapIntersect}
ranges(joined) <- ranges(pintersect(joined, subject[subjectHits(hits)]))
```

The typical aggregation is counting the number of hits overlapping a
query. In general, aggregation starts by grouping the subject hits by
query hits, which we express as a coercion to a *List*,
```{r hitsAsList}
hitsByQuery <- as(hits, "List")
```
The result is an *IntegerList*, a type of *AtomicList*. *AtomicList*
objects have many methods for efficient aggregation. In this case, we
just call `lengths()` to get the count:
```{r lengthsHits}
counts <- lengths(hitsByQuery)
```
Since this a common operation, there are shortcuts,
```{r countHits}
counts <- countQueryHits(hits)
```
or even shorter and more efficient,
```{r countOverlaps}
counts <- countOverlaps(query, subject, ignore.strand=TRUE)
unname(counts)
```

Often, we want to combine joins and aggregations. For example, we may
want to annotate each query with the maximum score among the subject
hits,
```{r joinMax}
query$maxScore <- max(extractList(subject$score, hitsByQuery))
subset(query, maxScore > 0)
```

In rare cases, we can more or less arbitrarily select one of the
subject hits. The `select=` argument to `findOverlaps()` automatically
selects an "arbitrary", "first" (in subject order) or "last" subject
range,
```{r select-first}
hits <- findOverlaps(query, subject, select="first", ignore.strand=TRUE)
hits <- findOverlaps(query, subject, select="arbitrary", ignore.strand=TRUE)
hits
```

## Exercises

1. Find the average intensity of the X and Y measurements for each
   each replicate over all positions in the query object
2. Add a new column to the intensities object that is the distance from
   each position to its closest gene (hint `IRanges::distance()`)
3. Find flanking regions downstream of the genes in gr that have width of 8bp
4. Are any of the intensities positions within the flanking region?
