---
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: false
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Interval operations for GenomicRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}  
---

# Basic interval operations for *GRanges* objects

There are many functions for manipulating *GRanges* objects. The
functions can be classified as *intra-range functions*, *inter-range
functions*, and *between-range functions*.

*Intra-range functions* operate on each element of a
*GRanges* object independent of the other ranges in the
object. For example, the `flank` function can be used to recover
regions flanking the set of ranges represented by the *GRanges*
object. So to get a *GRanges* object containing the ranges that
include the 10 bases upstream according to the direction of
"transcription" (indicated by the strand):
 
 ```{r flank}
g <- gr[1:3]
g <- append(g, gr[10])
flank(g, 10)
```

And to include the downstream bases:
 
```{r flank2}
flank(g, 10, start=FALSE)
```

A common use case for `flank()` is generating promoter regions based
on the transcript ranges. There is a convenience function that by
default generates a region starting 2000bp upstream and 200bp
downstream of the TSS,

```{r promoters}
promoters(g)
```

To ignore strand/transcription and assume the orientation of left to
right use `unstrand()`,
```{r unstrand}
flank(unstrand(g), 10)
```

Other examples of intra-range functions include `resize()` and
`shift()`. The `shift()` function will move the ranges by a specific number
of base pairs, and the `resize()` function will set a specific width, by
default fixing the "transcription" start (or just the start when
strand is "*"). The `fix=` argument controls whether the "start",
"end" or "center" is held constant.
 
```{r shiftAndResize}
shift(g, 5)
resize(g, 30)
```
 
The `r BiocStyle::Biocpkg("GenomicRanges")` help page `?"intra-range-methods"`
summarizes these methods.

*Inter-range functions* involve comparisons between ranges in a single
*GRanges* object and typically aggregate ranges. For instance, the
`reduce()` function will merge overlapping and adjacent ranges to
produce a minimal set of ranges representing the regions covered by
the original set.
 
```{r reduce}
reduce(gr)
reduce(gr, ignore.strand=TRUE)
```
 
Rarely, it useful to complement the (reduced) ranges. Note that the
universe is taken as the entire sequence span in all three strands (+,
-, *), which is often surprising when working with unstranded ranges.
```{r gaps}
gaps(g)
```
 
The `disjoin` function breaks up the ranges so that they do not
overlap but still cover the same regions:
 
```{r disjoin}
disjoin(g)
```
 
The `coverage` function counts how many ranges overlap each position
in the sequence universe of a *GRanges* object.
 
```{r coverage}
cov <- coverage(g)
cov[1:3]
```
The coverage is stored compactly as an *RleList*, with one *Rle*
vector per sequence. We can convert it to a *GRanges*,
```{r coverage_to_gr}
cov_gr <- GRanges(cov)
cov_gr
```
and even convert the *GRanges* form back to an *RleList* by computing
a weighted coverage,
```{r coverage_to_rle}
cov <- coverage(cov_gr, weight="score")
```

The *GRanges* derivative *GPos*, a compact representation of width 1
ranges, is useful for representing coverage, although it cannot yet
represent the coverage for the entire human genome (or any genome with
over ~ 2 billion bp).
```{r}
GPos(cov[1:3])
```

These inter-range functions all generate entirely new sets of
ranges. The return value is left unannotated, since there is no
obvious way to carry the metadata across the operation. The user is
left to map the metadata to the new ranges. Functions like `reduce()`
and `disjoin()` facilitate this by optionally including in the
returned metadata a one-to-many reverse mapping from the aggregate
ranges to input ranges. For example, to average the score over a
reduction,
```{r reduce-mapping}
rg <- reduce(gr, with.revmap=TRUE)
rg$score <- mean(extractList(gr$score, rg$revmap))
```

See the `r BiocStyle::Biocpkg("GenomicRanges")` help page
`?"inter-range-methods"` for additional help.

## Interval set operations for *GRanges* objects

*Between-range functions* calculate relationships between different
*GRanges* objects. Of central importance are
`findOverlaps` and related operations; these are discussed
below.  Additional operations treat *GRanges* as mathematical
sets of coordinates; `union(g, g2)` is the union of the
coordinates in `g` and `g2`. Here are examples for
calculating the `union`, the `intersect` and the
asymmetric difference (using `setdiff`).
 
```{r intervals1}
g2 <- head(gr, n=2)
union(g, g2)
intersect(g, g2)
setdiff(g, g2)
```

Related functions are available when the structure of the
*GRanges* objects are 'parallel' to one another, i.e., element
1 of object 1 is related to element 1 of object 2, and so on. These
operations all begin with a `p`, which is short for
parallel. The functions then perform element-wise, e.g., the union of
element 1 of object 1 with element 1 of object 2, etc. A requirement
for these operations is that the number of elements in each
*GRanges* object is the same, and that both of the objects have
the same seqnames and strand assignments throughout.
 
```{r intervals2}
g3 <- g[1:2]
ranges(g3[1]) <- IRanges(start=105, end=112)
punion(g2, g3)
pintersect(g2, g3)
psetdiff(g2, g3)
```

For more information on the `GRanges` classes be sure to consult
the manual page.
 
```{r manPage, eval=FALSE}
?GRanges
```
 
A relatively comprehensive list of available functions is discovered
with
 
```{r granges-methods, eval=FALSE}
methods(class="GRanges")
``` 
